### 2020-04-24 ncov.tex
minor changes
editing

\section*{Abstract}
A novel coronavirus (SARS-CoV-2) has recently emerged as a global threat. 
As the epidemic progresses, many disease modelers have focused on estimating the basic reproductive number \Ro -- the average number of secondary cases caused by a primary case in an otherwise susceptible population.
The modeling approaches and resulting estimates of \Ro vary widely, despite relying on similar data sources.
Here, we present a novel statistical framework for comparing and combining different estimates of \Ro across a wide range of models by decomposing the basic reproductive number into three key quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
We then apply our framework to early estimates of \Ro for the SARS-CoV-2 outbreak.
We show that many early \Ro estimates are overly confident.
Our results emphasize the importance of propagating uncertainties in all components of \Ro, including the shape of the generation-interval distribution, in efforts to estimate \Ro at the outset of an epidemic.

@@

\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{3.5cm}|p{2.5cm}|p{2.7cm}|l}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ & \\
\hline
Study 1 & 1.5--3.5 & 10 & 1 & Bedford \textit{et al.} \cite{bedfordncov} \\
\hline
Study 2 & 2.5 (1.5--3.5)$^\ast$ & 8.4 & unspecified$^\dagger$ & Imai \textit{et al.} \cite{imaincov} \\
\hline
Study 3 & 2.92 (95\% CI: 2.28--3.67) & 8.4 & 0.2 & Liu \textit{et al.} \cite{liuncov} \\
\hline
Study 4 & 3.8 (95\% CI: 3.6--4.0) & 7.6 & 0.5 & Read \textit{et al.} \cite{readncov} \\
\hline
Study 5 & 2.2 (90\% CI: 1.4--3.8) & 7--14 & 0.5 & Riou and Althaus \cite{riouncov} \\
\hline
Study 6 & 5.47 (95\% CI: 4.16--7.10)$^\ddagger$ & 7.6--8.4 & 0.2 & Zhao \textit{et al.} \cite{zhaoncov} \\
\hline
Study 7 & 2.0--3.1 & 6--10 & 0 & Majumder and Mandl \cite{majumderncov} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Reported estimates of the basic reproductive number and the assumptions about the generation-interval distributions.}
Estimates of \Ro and their assumptions about the shape of the generation interval distributions were collected from 7 studies.
$^\ast$We treat these intervals as a 95\% confidence interval in our analysis.
$^\dagger$We assume $\kappa = 0.5$ in our analysis.
$^\ddagger$The authors presented \Ro estimates under different assumptions regarding the reporting rate; we use their baseline scenario in our analysis to remain consistent with other studies, which do not account for changes in the reporting rate.
}
\end{table}

@@

Here, we use the gamma approximation framework \citep{park2019practical} to (i) characterize the amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution and (ii) assess the degree to which these uncertainties affect the estimate of \Ro.
Assuming that generation intervals follow a gamma distribution with the mean $\bar G$ and the squared coefficient of variation $\kappa$, we have
\begin{linenomath*}
\begin{equation}
\Ro = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
\end{linenomath*}
This equation demonstrates that a generation-interval distribution that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$) will give a higher estimate of \Ro for the same value of $r$ \citep{wallinga2007generation}.

@@

As most studies do not report their estimates of the exponential growth rate, we first recalculate the exponential growth rate that correspond to their model assumptions.
We do so by modeling reported distributions of the reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion parameter $\kappa$ with appropriate probability distributions;
we used gamma distributions to model values reported with confidence intervals and uniform distributions to model values reported with ranges.
For example, Study 3 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).
For each study $i$, we construct a family of parameter sets by drawing 100,000 random samples from the probability distributions (Table 2) that represent the estimates of \Rx{0i} and the assumed values of $\bar G_i$ and $\kappa_i$ and calculate the exponential growth rate $r_i$ via the inverse of \eref{gamma}:
\begin{linenomath*}
\begin{equation}
r_i = \frac{\Rx{0i}^{\kappa_i} - 1}{\kappa_i \bar{G}_i}.
\end{equation}
\end{linenomath*}
This allows us to approximate the probability distributions of the estimated exponential growth rates by each study;
uncertainties in the probability distributions that we calculate for the estimated exponential growth rates will reflect the methods and assumptions that the studies rely on.

@@

We run 4 independent Markov Chain Monte Carlo chains each consisting of 500,000 burnin steps and 500,000 sampling steps.
Posterior samples are thinned every 1000 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all hyperparameters \citep{gelman1992inference};
trace plots and marginal posterior distribution plots are presented in Appendix.
95\% confidence intervals are calculated by taking 2.5\% and 97.5\% quantiles from the marginal posterior distribution for each parameter.

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 23--26 (\fref{eff}A).
Comparing the models that include only some sources of uncertainty to the ``all'' model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (``GI mean''), which lacks growth-rate uncertainty, is relatively narrow.
In this case, propagating error from the mean generation interval has negligible effect compared to propagating the uncertainty in $r$.
Uncertainty in the generation-interval dispersion also has important effects as it determines the functional form of the relationship between $r$ and \Ro (compare ``growth rate + GI mean'' with ``all'').
For example, reducing the dispersion parameter $\kappa$ from 1 (assuming exponentially distributed generation intervals) to 0 (assuming fixed generation intervals) changes the $r$--\Ro\ relationship from linear to exponential, therefore increasing the sensitivity of \Ro estimates to $r$ and $\bar G$.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{compare_R0.pdf}
\caption{
\textbf{Sensitivity of the reported \Ro estimates with respect to our pooled estimates of the underlying parameters.}
We replace the reported parameter values (growth rate $r$, GI mean $\bar G$, and GI variation $\kappa$) with our corresponding pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$) one at a time and recalculate \Ro (\textbf{growth rate}, \textbf{GI mean}, and \textbf{GI variation}).
The pooled estimate of \Ro is calculated from the joint posterior distribution of $\mu_r$, $\mu_G$, and $\mu_\kappa$ (\textbf{all});
this corresponds to replacing all reported parameter values with our pooled estimates, which gives identical results across all studies.
Horizontal dashed lines represent the 95\% confidence intervals of our pooled estimate of \Ro.
The reported \Ro estimates (\textbf{base}) have been adjusted to show the approximate 95\% confidence interval using the probability distributions that we defined if they had relied on different measures for parameter uncertainties.
}
\label{fig:R0}
\end{figure}

>>>>

\section*{Abstract}
A novel coronavirus (SARS-CoV-2) has recently emerged as a global threat. 
As the epidemic progresses, many disease modelers have focused on estimating the basic reproductive number \Ro\ --- the average number of secondary cases caused by a primary case in an otherwise susceptible population.
The modeling approaches and resulting estimates of \Ro vary widely, despite relying on similar data sources.
Here, we present a novel statistical framework for comparing and combining different estimates of \Ro across a wide range of models by decomposing the basic reproductive number into three key quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
We then apply our framework to early estimates of \Ro for the SARS-CoV-2 outbreak.
We show that many early \Ro estimates are overly confident.
Our results emphasize the importance of propagating uncertainties in all components of \Ro, including the shape of the generation-interval distribution, in efforts to estimate \Ro at the outset of an epidemic.

@@

\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{3.5cm}|p{2.5cm}|p{2.7cm}|l}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ & \\
\hline
Study 1 & 1.5--3.5 & 10 & 1 & \cite{bedfordncov} \\
\hline
Study 2 & 2.5 (1.5--3.5)$^\ast$ & 8.4 & unspecified$^\dagger$ & \cite{imaincov} \\
\hline
Study 3 & 2.92 (95\% CI: 2.28--3.67) & 8.4 & 0.2 & \cite{liuncov} \\
\hline
Study 4 & 3.8 (95\% CI: 3.6--4.0) & 7.6 & 0.5 & \cite{readncov} \\
\hline
Study 5 & 2.2 (90\% CI: 1.4--3.8) & 7--14 & 0.5 & \cite{riouncov} \\
\hline
Study 6 & 5.47 (95\% CI: 4.16--7.10)$^\ddagger$ & 7.6--8.4 & 0.2 & \cite{zhaoncov} \\
\hline
Study 7 & 2.0--3.1 & 6--10 & 0 & \cite{majumderncov} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Reported estimates of the basic reproductive number and the assumptions about the generation-interval distributions.}
Estimates of \Ro and their assumptions about the shape of the generation interval distributions were collected from 7 studies.
$^\ast$We treat these intervals as a 95\% confidence interval in our analysis.
$^\dagger$We assume $\kappa = 0.5$ in our analysis.
$^\ddagger$The authors presented \Ro estimates under different assumptions regarding the reporting rate; we use their baseline scenario in our analysis to remain consistent with other studies, which do not account for changes in the reporting rate.
}
\end{table}

@@

Here, we use the gamma approximation framework \citep{park2019practical} to (i) characterize the amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution and (ii) assess the degree to which these uncertainties affect the estimate of \Ro.
Assuming that generation intervals follow a gamma distribution with mean generation interval $\bar G$ and generation-interval dispersion $\kappa$, represented by the squared coefficient of variation of a gamma distribution, we have
\begin{linenomath*}
\begin{equation}
\Ro = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
\end{linenomath*}
This equation demonstrates that a generation-interval distribution that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$) will give a higher estimate of \Ro for the same value of $r$ \citep{wallinga2007generation}.

@@

As most studies do not report their estimates of the exponential growth rate, we first recalculate the exponential growth rate that correspond to their model assumptions.
We do so by modeling reported distributions of the reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ with appropriate probability distributions;
we used gamma distributions to model values reported with confidence intervals and uniform distributions to model values reported with ranges.
For example, Study 3 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).
For each study $i$, we construct a family of parameter sets by drawing 100,000 random samples from the probability distributions (Table 2) that represent the estimates of \Rx{0i} and the assumed values of $\bar G_i$ and $\kappa_i$ and calculate the exponential growth rate $r_i$ via the inverse of \eref{gamma}:
\begin{linenomath*}
\begin{equation}
r_i = \frac{\Rx{0i}^{\kappa_i} - 1}{\kappa_i \bar{G}_i}.
\end{equation}
\end{linenomath*}
This allows us to approximate the probability distributions of the estimated exponential growth rates by each study;
uncertainties in the probability distributions that we calculate for the estimated exponential growth rates will reflect the methods and assumptions that the studies rely on.

@@

We run 4 independent Markov Chain Monte Carlo chains each consisting of 500,000 burnin steps and 500,000 sampling steps.
Posterior samples are thinned every 1000 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all hyperparameters \citep{gelman1992inference};
trace plots and marginal posterior distribution plots are presented in Appendix.
95\% credible intervals (CI) are calculated by taking 2.5\% and 97.5\% quantiles from the marginal posterior distribution for each parameter.

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 23--26 (\fref{eff}A).
Comparing the models that include only some sources of uncertainty to the ``all'' model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (``GI mean''), which lacks growth-rate uncertainty, is relatively narrow.
In this case, propagating error from the mean generation interval has negligible effect compared to propagating the uncertainty in $r$.
Uncertainty in the generation-interval dispersion $\kappa$ also has important effects as it determines the functional form of the relationship between $r$ and \Ro (compare ``growth rate + GI mean'' with ``all'').
For example, reducing $\kappa$ from 1 (assuming exponentially distributed generation intervals) to 0 (assuming fixed generation intervals) changes the $r$--\Ro\ relationship from linear to exponential, therefore increasing the sensitivity of \Ro estimates to $r$ and $\bar G$.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{compare_R0.pdf}
\caption{
\textbf{Sensitivity of the reported \Ro estimates with respect to our pooled estimates of the underlying parameters.}
We replace the reported parameter values (growth rate $r$, mean generation interval $\bar G$, and generation-interval dispersion $\kappa$) with our corresponding pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$) one at a time and recalculate \Ro (\textbf{growth rate}, \textbf{GI mean}, and \textbf{GI variation}).
The pooled estimate of \Ro is calculated from the joint posterior distribution of $\mu_r$, $\mu_G$, and $\mu_\kappa$ (\textbf{all});
this corresponds to replacing all reported parameter values with our pooled estimates, which gives identical results across all studies.
Horizontal dashed lines represent the 95\% confidence intervals of our pooled estimate of \Ro.
The reported \Ro estimates (\textbf{base}) have been adjusted to show the approximate 95\% confidence interval using the probability distributions that we defined if they had relied on different measures for parameter uncertainties.
}
\label{fig:R0}
\end{figure}
